---
title: "Reproducible Research Peer Assesment 1"
output:
  html_document:
    fig_caption: yes
    fig_height: 4.5
    fig_width: 6.5
    keep_md: yes
---
##Introduction
Personal activity monitoring devices allow users to collect a large amount of data about themselves. In this report, we will analyze of set of such data.  
The data used in this report was obtained [here](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) (link is to zip file).

Setup steps and packages used:
```{r setup, echo=TRUE, results='hide', message=FALSE, warning=FALSE, cache=TRUE}
setwd("~/data")
rm(list = ls())
library(dplyr)
library(ggplot2)
library(reshape2)
library(gridExtra)
```


##Load and Process Data 
First, we need to load and read in the data. We can see that the data consists of three variables, "steps", "date", and "interval".  

* Steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)  
* date: The date on which the measurement was taken in YYYY-MM-DD format  
* interval: Identifier for the 5-minute interval in which measurement was taken  

In addition, we will do a few pre-processing steps -- which will create dataframes that summarize the daily steps by date, and also by interval (time of day).

```{r processing, echo=TRUE, dependson=1, cache=TRUE}
fileurl="http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
filename <- "activity.zip"
if (file.exists(filename)) {
  print("File present!")
  } else {download.file(fileurl, destfile="activity.zip")}
#This is the main data
activity <- read.csv(unz("activity.zip","activity.csv"))
head(activity) 
#For now, remove NA's
act_clean <- activity[!is.na(activity$steps),]
#Agrregated dataframes by day and by interval
bydate <- act_clean %>%
          group_by(date) %>%
          summarize(steps=sum(steps))
byint <- act_clean %>%
          group_by(interval) %>%
          summarize(steps=mean(steps))
```

##What is mean total number of steps taken per day? 
Using a simple histogram, we get an idea of the average daily total steps.  

```{r plot1, echo=TRUE, dependson=c(1,2), cache=TRUE, results='hold'}
means <- data.frame(Mean=signif(mean(bydate$steps), digits=7), Median=signif(median(bydate$steps), digits=7))
plot <- ggplot(data=bydate, aes(bydate$steps))
plot1 <- plot + geom_histogram(stat="bin", binwidth=max(bydate$steps)/10,color="blue", fill="white")
plot1 <- plot1 + 
  labs(title="Total number of steps taken per day", x="Total steps per day", y="Count of summed steps") + 
  geom_vline(xintercept=mean(bydate$steps), color="blue", lty=2) + 
  geom_vline(xintercept=median(bydate$steps),color="red", lty=5) + 
  annotation_custom(tableGrob(means), xmin=15000, xmax=25000, ymin=8, ymax=15)
plot1

```

##What is the average daily activity pattern?
Earlier, a dataframe reporting average steps by each 5-min interval was created, so we can use that dataframe to make a line plot of activity at each 5-minute interval, averaged over all days. In the dataset, the 5-minute intervals are identified by an index from 0 to 2355.

```{r plot2, echo=TRUE, dependson=1, cache=TRUE}
#calculate maximum avg. steps and its matching interval
maxint<-byint$interval[which.max(byint$steps)]
maxstep<-max(byint$steps)
maxes <- data.frame(MaxInterval=maxint, MaxSteps=round(maxstep))
plot2 <- ggplot(data=byint, aes(x=interval, y=steps))
plot2 <- plot2 + geom_line() +
  labs(title="Average Daily Activity\n in 5-minute intervals",
     x="Interval",
     y="Average Steps Taken") +
  geom_vline(xintercept=maxint) +
  annotation_custom(tableGrob(maxes), xmin=1500, xmax=2000, ymin=125, ymax=150)
plot2
```


##Imputing missing values
The original dataset had a missing values, signified by "NA", which we initially ignored. First we figure out how many missing values there are.

There are `r nrow(activity)-nrow(act_clean)` missing values, and `r nrow(act_clean)` complete cases in the original dataset. To fill in this data, we can use the average values from each 5-minute interval. We could have also used the daily average, but since the previous plot showed that there is a lot of variation by time of day, it seems "safer" to use the average at each 5-min interval.

To do this, we go back to the original dataset. We also use the interval means dataset (`byint`) created for the second plot, which provides us the mean interval values to substitute in at each missing datapoint.

```{r plot3, dependson=2, cache=TRUE}
act2<-activity
for (i in 1:nrow(act2)) {
    if (is.na(act2$steps[i])) {
        act2$steps[i] <- byint[which(act2$interval[i] == byint$interval), ]$steps
    }
}
sum(is.na(act2)) #check that there's no more NAs
head(act2)
```

**Did the data change by our "impute" procedure?**
To look at the effect of [fill in]

```{r plot4, dependson=c(2,5), cache=TRUE}
bydate2 <- act2 %>% 
        group_by(date) %>%
        summarize(steps=sum(steps)) 
plot <- ggplot(data=bydate2, aes(bydate2$steps))
plot4 <- plot + geom_histogram(stat="bin",
                               binwidth=max(bydate$steps)/10,
                               color="blue", 
                               fill="white")
plot4 <- plot4 + labs(title="Total number of steps taken per day",
                      x="Total steps per day",
                      y="Count of summed steps")
plot4

means <- data.frame(Mean=mean(bydate$steps), Median=median(bydate$steps), Mean_new=mean(bydate2$steps), Median_new=median(bydate2$steps))
print(means)
```

The mean and median have not changed much by "imputing" the missing data. The main effect of the imputation is that the frequencies at each histogram bin has increased. 

##Are there differences in activity patterns between weekdays and weekends?

```{r plot5, dependson=c(2,5)}
#change date format in dataset
act2$date <- as.Date(strptime(act2$date, format="%Y-%m-%d"))
#create a "day of week" variable
act2$dow <- weekdays(act2$date)
#Subset the data into Weekend and Weekday, and replace day of week 
wkend <- act2 %>% subset(dow %in% c("Saturday", "Sunday")) %>% mutate(dow2="Weekend")
wkday <- act2 %>% 
        subset(!dow %in% c("Saturday", "Sunday")) %>%
        mutate(dow2="Weekday")
#combine the two sets
act3 <- rbind(wkday, wkend)    
#get aggregate dataset for plot
act4 <- act3 %>%
        group_by(interval, dow2) %>%
        summarize(steps=mean(steps))
plot <- ggplot(act4, aes(x=interval, y=steps)) + 
  geom_line(color="dark green", type="l") + 
  facet_grid(dow2~., scales="free", space="free")
plot <- plot + 
  labs(x="Interval", 
       y="Mean Steps", 
       title="Mean Steps by 5-min Interval\n Weekend vs. Weekdays") 
print(plot)
```

From this plot, we see that our anonymous walker walks more during weekends, but each weekday, there is a lot of walking right near the interval at 830 (which we sort of knew from before). The final graph gives us a clue as to the overall walking pattern: More walking on weekends, but on weekdays, our person does a lot of walking around ~8:30AM. This pattern fits with a person who maybe walks to work everyday, and then takes a lot of walks on weekend days. 



