---
title: "Reproducible Research Peer Assesment 1"
output:
  html_document:
    fig_caption: yes
    fig_height: 6
    fig_width: 6
    keep_md: yes
---

##Introduction
Personal activity monitoring devices allow users to collect a large amount of data about themselves. In this report, we will analyze of set of such data.  
The data used in this report was obtained [here](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip).

Setup steps and packages used:
```{r initial, results='hide', message=FALSE, warning=FALSE}
setwd("~/data")
rm(list = ls())
library(dplyr)
library(ggplot2)
library(scales)
library(gridExtra)
library(lubridate)
library(knitr)
opts_chunk$set(cache=TRUE, echo=TRUE)
```


##Load and Process Data 
First, we need to load and read in the data. We can see that the data consists of three variables, "steps", "date", and "interval".  

* Steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)  
* date: The date on which the measurement was taken in YYYY-MM-DD format.
+The data come from a two-month period of data.    
* interval: Identifier for the 5-minute interval in which measurement was taken.
+This interval is given as a numeric value from 0 to 2355, representing five-minute segments (i.e 830=8:30 AM, while 2200=10 PM).  

The pre-processing steps will create dataframes that summarize the daily steps by date, and also by interval, in a subset of the original data that has NAs removed. 

```{r setup}
fileurl="http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
filename <- "activity.zip"
if (file.exists(filename)) {
  print("File present!")
  } else {download.file(fileurl, destfile="activity.zip")}
#This is the main data
activity <- read.csv(unz("activity.zip","activity.csv"))
head(activity) 
#For now, remove NA's
act_clean <- activity[!is.na(activity$steps),]
#Agrregated dataframes by day and by interval
bydate <- act_clean %>%
          group_by(date) %>%
          summarize(steps=sum(steps)) 
      
byint <- act_clean %>%
          group_by(interval) %>%
          summarize(steps=mean(steps))
#creating a as.POSIXct interval datetime variable and adding a dummy YMD to the intervals
byint_plot <- byint 
byint_plot$timeint<-parse_date_time(paste("2012-10-01",sprintf("%04d",byint_plot$interval)), "Y!m*!d!H!M!")
```

##What is mean total number of steps taken per day?
To answer the question, we plot a histogram of the dataset bydate, which contains the summed steps taken for every day of the dataset. Using the  pacakge [gridExtra](http://cran.r-project.org/web/packages/gridExtra/gridExtra.pdf), we can annotate the plot with the calculated mean and median. 

```{r plot1, echo=TRUE, dependson="setup", fig.cap="Distribution of total daily steps"}
table1 <- data.frame(Mean=sprintf("%1.3f",mean(bydate$steps)), 
                     Median=sprintf("%1.3f", median(bydate$steps)))

plot <- ggplot(data=bydate, aes(bydate$steps))
plot1 <- plot + geom_histogram(stat="bin", binwidth=max(bydate$steps)/15,color="dark green", fill="white")
plot1 <- plot1 + 
  labs(title="Total number of steps taken per day", x="Total steps per day", y="Frequency") + 
  geom_vline(xintercept=mean(bydate$steps), color="blue", lty=2) + 
  geom_vline(xintercept=median(bydate$steps),color="red", lty=5) 
grid.arrange(plot1, tableGrob(table1), nrow=2, heights=c(1,0.1))
```

##What is the average daily activity pattern?
The dataframe byint contains the average steps at each 5-min interval. Each interval is averaged across all days reported in the data (currently with NAs removed). We can use that dataframe to make a line plot and annotate the plot with the maximum average steps taken and the 5-minute interval in which it happened. 


```{r plot2, echo=TRUE, dependson="setup", cache=TRUE, fig.cap="Daily Activity Pattern"}
#calculate maximum avg. steps and its matching interval
maxint<-byint$interval[which.max(byint$steps)]
maxstep<-max(byint_plot$steps)
table2 <- data.frame(Most_Active_Interval=maxint, Steps=round(maxstep))

plot2 <- ggplot(data=byint_plot, aes(x=timeint, y=steps))
plot2 <- plot2 + geom_line(color="red") +
  labs(title="Average Daily Activity in 5-minute intervals",
     x="Time of Day (HH:MM)",
     y="Average Steps Taken") + 
  scale_x_datetime(labels=date_format("%H:%M"))
grid.arrange(plot2, tableGrob(table2), nrow=2, heights=c(1,0.1))
```


##Imputing missing values
The original dataset had missing values, signified by "NA", which we initially ignored. 
The antijoin function of the dplyr package can compare two datasets: [it returns all values in x that are not matched in y, and only keeps columns from x.](https://stat545-ubc.github.io/bit001_dplyr-cheatsheet.html). This will tell us which entries corresponded to the NA values. 

```{r missingdata, dependson="setup"}
#return na rows with anti_join
bydate_missing <- anti_join(activity, act_clean, by="steps")
days <-distinct(bydate_missing, Missing=as.Date(bydate_missing$date))
#Information on missing Dates
days$Missing
data.frame(TotalObs=nrow(activity), CompleteObs=nrow(act_clean), MissingObs=nrow(bydate_missing))
```

There are `r nrow(bydate_missing)` missing rows to impute out of a total `r nrow(activity)`, representing time intervals missing data. 
To fill in these missing time intervals, we can use the average values from each 5-minute interval that was calculated earlier and stored in byint. Essentially, we will assume that on the 8 missing days, our anonymous "Quantified Self"-er had an activity pattern close to the average activity pattern over this period. 

```{r imputing, dependson="setup", cache=TRUE}
act2<-activity
for (i in 1:nrow(act2)) {
    if (is.na(act2$steps[i])) {
        act2$steps[i] <- byint[which(act2$interval[i] == byint$interval), ]$steps
    }
}
sum(is.na(act2)) #check that there's no more NAs
head(act2) 
```

**How did imputing data affect the mean and median?**
The original dataset had data missing mostly from 8 sporadic days out of the 61 day reporting period. These missing data were imputed with a mean value, so we expect that the overall mean should not be changed much. To test this, we repeat the same process we applied to the original dataset to get a histogram with calculated mean and median. 

```{r plot3, dependson=c("setup", "plot1", "imputing"), fig.cap="Before and After Imputing Missing Data"}
bydate2 <- act2 %>% 
        group_by(date) %>%
        summarize(steps=sum(steps)) 
tableN <- data.frame(NewMean=sprintf("%1.3f",mean(bydate2$steps)), NewMedian=sprintf("%1.3f",median(bydate2$steps)))
table3<-cbind(table1, tableN)
                    
plot3 <- ggplot(data=bydate2, aes(steps))
plot3 <- plot3 + geom_histogram(stat="bin", binwidth=max(bydate2$steps)/15,color="dark green", fill="white") + ylim(0,22)
plot3 <- plot3 + 
  labs(title="Total number of steps taken per day (Imputed Data)", x="Total steps per day", y="Frequency") + 
  geom_vline(xintercept=mean(bydate2$steps), color="blue", lty=2) + 
  geom_vline(xintercept=median(bydate2$steps),color="red", lty=5) 
#place new histogram next to original histogram, and with same y-axis scale
plot1 <- plot1 + ylim(0,22)
grid.arrange(plot3, plot1, tableGrob(table3), nrow=3, heights=c(1,1,0.25))
```

We can confirm that the mean and median have not changed much in the imputed data. After imputation, the mean and median are closer to each other. We expected that the mean should not change since the average daily pattern was used to impute. The histogram shows that the main effect was to increase the number of counts in the bin that contains the mean and median. 

##Are there differences in activity patterns between weekdays and weekends?

```{r plot4, dependson=c(2,3,6), fig.cap="Daily activity on Weekdays vs. Weekends"}
#change date format in dataset
act2$date <- as.Date(strptime(act2$date, format="%Y-%m-%d"))
#create a "day of week" variable
act2$dow <- weekdays(act2$date)
#Subset the data into Weekend and Weekday, and replace day of week 
wkend <- act2 %>%
        subset(dow %in% c("Saturday", "Sunday")) %>%
        mutate(dow2="Weekend")
wkday <- act2 %>% 
        subset(!dow %in% c("Saturday", "Sunday")) %>%
        mutate(dow2="Weekday")
#combine the two sets
act3 <- rbind(wkday, wkend)    
#get aggregate dataset for plot
act4 <- act3 %>%
        group_by(interval, dow2) %>%
        summarize(steps=mean(steps))
act4plot<-act4
act4plot$timeint<-parse_date_time(paste("2012-10-01",sprintf("%04d",act4$interval)), "Y!m*!d!H!M!")
plot4 <- ggplot(data=act4plot, aes(x=timeint, y=steps)) + 
  geom_line(color="dark green", type="l", aes(group=dow2)) + 
  facet_grid(dow2~., scales="free", space="free")
plot4 <- plot4 + 
  labs(x="Time of Day (HH:MM)", 
       y="Mean Steps", 
       title="Mean Steps by 5-min Interval: Weekend vs. Weekdays") +
  scale_x_datetime(labels=date_format("%H:%M"))
plot4
```

From this plot, we see that on weekends, there are more intervals where the number of steps is >100. On weekdays, there is a peak of activity in the mornings, around 8-9AM. So the overall activity on weekends seems to be higher but the weekdays have the biggest 5-min peak, at 8:30. This pattern fits with a person who takes a walk every weekday morning, or walks to work. This person's weekends are fairly active, and they seem to spend a greater part of the weekend days moving around. 

