---
title: "Reproducible Research Peer Assesment 1"
output:
  html_document:
    fig_caption: yes
    keep_md: yes
---

##Introduction  

  Personal activity monitoring devices, like "[fitness bands](https://www.google.com/search?q=fitness+bands)" allow users to collect a large amount of data about their own physical activity, using measures like number of steps. Sometimes, the fitness device comes with software for basic analyses.  
  This report provides methods for processing and analyzing such fitness data, and may be beneficial as an example for people who find the commercial pre-packagd analyses limited or wish to connect their data to their own personal applications. The data used in this report was obtained [here](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip).

Setup steps and packages used:
```{r initial, results='hide', message=FALSE, warning=FALSE}
setwd("~/data")
rm(list = ls())
library(dplyr)
library(ggplot2) 
library(scales)
library(gridExtra)
library(lubridate)
library(knitr)
opts_chunk$set(cache=TRUE, echo=TRUE)
```


##Load and Process Data  

  The data consists the following variables:  

* Steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)  
* date: The date on which the measurement was taken in YYYY-MM-DD format.
* interval: Identifier for the 5-minute interval in which measurement was taken.
  + This interval is given as a numeric value from 0 to 2355, representing five-minute segments (i.e 830=8:30 AM, while 2200=10 PM).  

  We first create summary datasets of the total steps per day, and the daily activity pattern: `bydate`, which is the summed steps by day and `byint`, which contains the average steps at each 5-min interval, averaged over all days. Additionally, a formatted datetime variable is created. 

```{r setup}
fileurl="http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
filename <- "activity.zip"
if (!file.exists(filename)) {
    download.file(fileurl, destfile="activity.zip")} 
#This is the main data
activity <- read.csv(unz("activity.zip","activity.csv"))
head(activity) 
#For now, remove NA's
act_clean <- activity[!is.na(activity$steps),]
#Agrregated dataframes by day and by interval
bydate <- act_clean %>%
          group_by(date) %>%
          summarize(steps=sum(steps)) 
byint <- act_clean %>%
          group_by(interval) %>%
          summarize(steps=mean(steps))
#format the datetime values 
bydate$date <- as.Date(strptime(bydate$date, format="%Y-%m-%d"))
byint$timeint<-as.POSIXct(sprintf("%04d",byint$interval), tz="UTC", format="%H%M")
```

##What is mean total number of steps taken per day?  
  We make a histogram of `bydate`, and a summary table of mean and median (to at least 2 decimal places) using [gridExtra](http://cran.r-project.org/web/packages/gridExtra/gridExtra.pdf).

```{r plot1, echo=TRUE, dependson="setup", fig.cap="Distribution of total daily steps"}
#calculate mean and median and put in a table
values <- c(mean(bydate$steps), median(bydate$steps))
table1 <- data.frame(values, row.names=c("Mean", "Median")) 
#plot
plot <- ggplot(data=bydate, aes(bydate$steps))
plot1 <- plot + geom_histogram(stat="bin", binwidth=max(bydate$steps)/15, color="black", fill="light green")
plot1 <- plot1 + 
  labs(title="Total number of steps taken per day", x="Total steps per day", y="Frequency") + 
  geom_vline(xintercept=mean(bydate$steps), color="blue", lty=2) + 
  geom_vline(xintercept=median(bydate$steps),color="red", lty=5) +
  theme_bw()
#layout of plot and table
grid.arrange(plot1, tableGrob(format(table1, nsmall=2), show.colnames = FALSE), nrow=2, heights=c(1,0.25))
```

##What is the average daily activity pattern?

  The dataframe `byint` contains the average steps at each 5-min interval. Each interval is averaged across all days reported in the data (currently with NAs removed). In the next figure, we plot the average steps against the time interval. The table reports the maximum steps and the associated interval (rounded). 

```{r plot2, echo=TRUE, dependson="setup", cache=TRUE, fig.cap="Daily Activity Pattern"}
#calculate maximum avg. steps and its matching interval
values <- rbind(byint$interval[which.max(byint$steps)], max(byint$steps))
table2 <- data.frame(values, row.names=c("Max Interval", "Steps"))
#plot
plot2 <- ggplot(data=byint, aes(x=timeint, y=steps))
plot2 <- plot2 + geom_line(color="black") +
  labs(title="Average Daily Activity in 5-minute intervals",
     x="Time of Day",
     y="Average Steps Taken") + 
  scale_x_datetime(breaks=date_breaks("5 hours"), minor_breaks=date_breaks("30 min"),
                   labels=date_format("%I:%M%p")) +
  theme_bw() 
#layout of plot and table
grid.arrange(plot2, tableGrob(round(table2), show.colnames=FALSE), nrow=2, heights=c(1,0.1))
```


##Imputing missing values  

  The original dataset had missing values, signified by "NA", which we initially ignored. The dplyr package can compare two datasets using the join function. We spefically use `anti_join()`: [it returns all values in x that are not matched in y, and only keeps columns from x.](https://stat545-ubc.github.io/bit001_dplyr-cheatsheet.html). The joined dataset will only contain NA entries, so that we can see which dates contain missing data. Also, the row count can be compared against the original dataset. 

```{r missingdata, dependson="setup"}
#return na rows with anti_join
missing <- anti_join(activity, act_clean, by="steps")
days <-distinct(missing, Missing=as.Date(missing$date))
#Information on missing Dates
days$Missing
data.frame(TotalObs=nrow(activity), CompleteObs=nrow(act_clean), MissingObs=nrow(missing))
```

  There are `r nrow(missing)` missing rows to impute out of a total `r nrow(activity)`, and these missing rows are associated with 8 days out of the 2-month dataset. We replace all the values of `steps` in `missing` with the corresponding mean value from `byint`. If we had a much larger dataset this solution may need to be changed to something faster. 
    
```{r imputing, dependson="setup", cache=TRUE}
for (i in 1:nrow(missing)) {
  missing$steps[i] = byint[which(missing$interval[i] == byint$interval), ]$steps
            }
act2 <- rbind(act_clean, missing)
sum(is.na(act2)) #check that there's no more NAs
```

**How did imputing data affect the mean and median?**  
  Since we used a mean value to replace missing data, we expect that the overall mean should not be changed much. We can test this by looking at the histogram and the calculated mean and median.

```{r plot3, dependson=c("setup", "plot1", "imputing"), fig.cap="Before and After Imputing Missing Data", fig.height=8, fig.width=8}
#aggregate new, imputed data
bydate2 <- act2 %>% 
        group_by(date) %>%
        summarize(steps=sum(steps)) 
#create summary table
values <- rbind(mean(bydate2$steps),median(bydate2$steps))
tableA <- data.frame(values, row.names=c("New Mean", "New Median"))
table3 <- rbind(table1, tableA)
#Combne the two sets of data for plotting
bydate <- mutate(bydate, Dataset="Original")
bydate2 <- mutate(bydate2, Dataset="Imputed")
bydate3 <- rbind(bydate, bydate2)
#plot
plot3 <- ggplot(data=bydate3, aes(x=steps))
plot3 <- plot3 + geom_histogram(stat="bin", binwidth=max(bydate2$steps)/15,color="dark green", fill="grey") + facet_wrap(~ Dataset) 
plot3 <- plot3 + labs(title="Total number of steps taken per day", x="Total steps per day", y="Frequency") + geom_vline(xintercept=mean(bydate2$steps), color="blue", lty=2) + geom_vline(xintercept=median(bydate2$steps),color="red", lty=5) + theme_bw()
#Plot and Table Layout
grid.arrange(plot3, tableGrob(format(table3,nsmall=2), show.colnames=FALSE), nrow=2, heights=c(1,0.25))
```

##Are there differences in activity patterns between weekdays and weekends?

```{r plot4, dependson=c(2,3,6), fig.cap="Daily activity on Weekdays vs. Weekends"}
#change date format in dataset
act2$date <- as.Date(strptime(act2$date, format="%Y-%m-%d"))
#create a "day of week" variable
act2$dow <- wday(act2$date, label=TRUE)
#Subset the data into Weekend and Weekday, and replace day of week 
wkend <- act2 %>%
        subset(dow %in% c("Sat", "Sun")) %>%
        mutate(dow2="Weekend")
wkday <- act2 %>% 
        subset(!dow %in% c("Sat", "Sun")) %>%
        mutate(dow2="Weekday")
#combine the two sets
act3 <- rbind(wkday, wkend)    
#get aggregate dataset for plot
act4 <- act3 %>%
        group_by(interval, dow2) %>%
        summarize(steps=mean(steps))
act4plot<-act4
act4plot$timeint<-parse_date_time(paste("2012-10-01",sprintf("%04d",act4$interval)), "Y!m*!d!H!M!")
plot4 <- ggplot(data=act4plot, aes(x=timeint, y=steps)) + 
  geom_line(color="dark blue", type="l", aes(group=dow2)) + 
  facet_grid(dow2~., scales="free", space="free")
plot4 <- plot4 + 
  labs(x="Time of Day", 
       y="Mean Steps", 
       title="Mean Steps by 5-min Interval: Weekend vs. Weekdays") +
  scale_x_datetime(breaks=date_breaks("5 hours"), minor_breaks=date_breaks("30 min"),
                   labels=date_format("%I:%M%p")) +
  theme_bw()
plot4
```


  From this plot, we see that on weekends, there are more intervals where the number of steps is >100, which implies that the weekends contain more sustained activity over the day than weekdays. On weekdays, there is a peak of activity in the mornings, around 8-9AM. This pattern fits with a person who takes a walk every weekday morning, or walks to work, and then engages in a lot of activity over the weekend.

##Conclusions  
  Using a downloaded .csv file of activity data as the input we can use R, Markdown and Knitr to create a customized analysis output from any giving dataset, which we can then use for several applications. For example, one can generate reports of activity levels as a way to track oneself to a healthier lifestyle. 
  
##Appendix: Additional plots which may be of interest  
  Here's two additional plots that I also made while playing around with this data. In the first one, I plotted the daily steps against the date, and made a very crude attempt at finding a trend in the data. The second plot is just an "expanded" version of the weekend vs. weekday plot.  

```{r extra, dependson=c(2,3,6,7,8), fig.cap="Additional plots", fig.height=8, fig.width=8}
#aggregate data by date and day of week
act5 <- act3 %>%
  group_by(date,dow) %>%
  summarize(steps=sum(steps))
#plot
plotX <- ggplot(act5, aes(x=ymd(date),y=steps)) + 
  geom_point(size=3) + 
  geom_hline(yintercept=mean(act5$steps), color="red", lty=1, lwd=2, alpha=0.5) 
plotX <- plotX + labs(title="Total Daily Steps\n From Oct 1 to Dec 1, 2012", 
                      x="Date", y="Total Daily Steps") +
  theme_bw()

#aggregate by interval and day
act6 <- act3 %>%
        group_by(interval, dow) %>%
        summarize(steps=mean(steps))
act6plot<-act6
act6plot$timeint<-parse_date_time(paste("2012-10-01",sprintf("%04d",act6plot$interval)), "Y!m*!d!H!M!")
plotY <- ggplot(data=act6plot, aes(x=timeint, y=steps)) + 
  geom_line(color="dark blue", type="l", aes(group=dow)) + 
  facet_wrap(~ dow)
plotY <- plotY + 
  labs(x="Time of Day", 
       y="Mean Steps", 
       title="Mean Steps by 5-min Interval\nand by Day of the Week") +
  scale_x_datetime(breaks=date_breaks("8 hours"), minor_breaks=date_breaks("1 hour"),
                   labels=date_format("%I:%M%p")) +
  theme_bw()
grid.arrange(plotX, plotY, nrow=2)
```

